Bayesian Deep Gaussian Processes for Correlated Functional Data: \\ A Case Study in Cosmological Power Spectra

### 2025-05-09 paper/main.tex
updates to section 2-3

Notice, in the center panel of Figure \ref{fig:plot_data}, the precision of each
low/high resolution spectra increases as $k$ increases.  The curves get progressively 
tighter from left to right.  We would like to conveniently wrap this domain-specific 
knowledge inside our Bayesian framework later.  To this end, \cite{moran2023mira} 
used the multiple low- and high-resolution spectra across all cosmologies to
obtained precision estimates, $p_1,\dots, p_n$, across the $n$ 
wavenumber values using a log-log regression model.
% and a multiplier $c$ for the increase in precision from the 
% low- to high- resolution output.
We use these estimates to define three diagonal precision matrices, 
one for each data type (perturbation theory, low resolution, and high resolution).
Specifically, denote $\Lambda_p$, $\Lambda_\ell$, and $\Lambda_h$ as the
diagonal matrices with elements
\[
\Lambda_p^{(ii)} = \begin{cases}
    10^5 &\text{for } k_i < 0.04 \\
    0  &\text{otherwise}\\
    \end{cases}
\quad
\Lambda_\ell^{(ii)} = \begin{cases}
    16p_i &\text{for } 0.04 \leq k_i < 0.25 \\
    0  &\text{otherwise}\\
    \end{cases}
\quad
\Lambda_h^{(ii)} = \begin{cases}
    cp_i &\text{for } 0.04 \leq k_i < 5 \\
    0  &\text{otherwise}\\
    \end{cases}
\]
for $i=1,\dots, n$.  In $\Lambda_p$, we use the sufficiently high precision 
of $10^5$ to impose the deterministic nature of the perturbation theory output.  
\textcolor{red}{(@Steve - what is $c$
exactly?)}  We then calculate a weighted average for each cosmology: 
$\bar y = \Lambda^{-1}(\Lambda_p y_p + \Lambda_h y_h + \Lambda_{\ell} \bar{y}_\ell)$, 
where $\Lambda = \Lambda_p + \Lambda_\ell + \Lambda_h$ and 
$\bar{y}_\ell = \frac{1}{16}\sum_{r=1}^{16} y_{\ell_r}$. In Figure \ref{fig:plot_data},
the weighted average is shown in dashed orange.  As an estimate of the true
underlying spectra, the weighted average fluctuates too drastically due 
to the limited number of model runs.  It fails to account for the
spatial dependence inherent in the smooth but stochastic spectra.
To account for these modeling challenges, we need a more complex model.

@@

Here we describe our first contribution: the use of a Bayesian hierarchical model 
to estimate the underlying spectra and quantify uncertainty for a particular cosmology,
independent of the others.  Let $D_n = \{X_n, \bar{Y}, \Lambda\}$ represent the aggregated 
data for a particular cosmology.  Let $S$ represent the true (but unknown) underlying
power matter spectra from which $D_n$ was generated.
\textcolor{red}{(What is X?  Why is $\bar{Y}$ capital here, but it wasn't earlier?
Since $\Lambda$ is fixed and doesn't change, I do not think we need to consider it part
of the data.)}

@@

Here, we introduce the Bayesian Gaussian Process (GP) model We assume 
the matter power spectrum $S$ for a given cosmology follows a GP with mean $\mu$ 
and covariance matrix $\Sigma_S$.

@@

We transform $X$ values to fall within $[0,1]$ and $Y$ so that the mean and variance 
are 0 and 1, respectively. This allows for a straightforward specification of priors, 
where the hyperparameters for the Mat\'ern covariance $\Sigma_S$ each take a vague prior. 
Specifically:

@@

\begin{align}
\bar Y|S &\sim \mathcal{N}_n(S,\Sigma_\varepsilon) \\
S &\sim GP\left(\mu, \Sigma_S(x,x')\right)
\end{align}

Upon integrating out $S$, we obtain $\bar Y \sim \mathcal{N}_n(\mu, \Sigma_S+\Sigma_\varepsilon)$. 
Additionally, this model yields the posterior distribution $S|\bar Y \sim \mathcal{N}_n(m, C)$, 
where $C=\left(\Sigma_S^{-1}+\Sigma_\varepsilon^{-1}\right)^{-1}$ and 
$m=C\left(\Sigma_\varepsilon^{-1}\bar Y+\Sigma_S^{-1}\mu\right)$. 
This distribution provides the posterior mean and corresponding uncertainty for 
$S$ within a given cosmology. 
% With this hierarchical model, we improve the efficiency of estimating $S|\bar{Y}$.

>>>>

Notice, in the center panel of Figure \ref{fig:plot_data}, the precision of each
low/high resolution spectra increases as $k$ increases.  The curves get progressively 
tighter from left to right.  We would like to conveniently wrap this domain-specific 
knowledge inside our Bayesian framework later.  To this end, \cite{moran2023mira} 
used the multiple low- and high-resolution spectra across all cosmologies to
obtain precision estimates, $p_1,\dots, p_n$, across the $n$ 
wavenumber values using a log-log regression model. A multiplier $c$ for 
the increase in precision from the low- to high- resolution output is also 
estimated with this regression model. We use these estimates to define three 
diagonal precision matrices, one for each data type (perturbation theory, 
low resolution average, and high resolution). Specifically, denote $\Lambda_p$, 
$\Lambda_\ell$, and $\Lambda_h$ as the diagonal matrices with elements
\[
\Lambda_p^{(ii)} = \begin{cases}
    10^5 &\text{for } k_i < 0.04 \\
    0  &\text{otherwise}\\
    \end{cases}
\quad
\Lambda_\ell^{(ii)} = \begin{cases}
    16p_i &\text{for } 0.04 \leq k_i < 0.25 \\
    0  &\text{otherwise}\\
    \end{cases}
\quad
\Lambda_h^{(ii)} = \begin{cases}
    cp_i &\text{for } 0.04 \leq k_i < 5 \\
    0  &\text{otherwise}\\
    \end{cases}
\]
for $i=1,\dots, n$.  In $\Lambda_p$, we use the sufficiently high precision 
of $10^5$ to impose the deterministic nature of the perturbation theory output.  
We then calculate a weighted average for each cosmology: 
$\bar y = \Lambda^{-1}(\Lambda_p y_p + \Lambda_h y_h + \Lambda_{\ell} \bar{y}_\ell)$, 
where $\Lambda = \Lambda_p + \Lambda_\ell + \Lambda_h$ and 
$\bar{y}_\ell = \frac{1}{16}\sum_{r=1}^{16} y_{\ell_r}$. In Figure \ref{fig:plot_data},
the weighted average is shown in dashed orange.  As an estimate of the true
underlying spectra, the weighted average fluctuates too drastically due 
to the limited number of model runs.  It fails to account for the
spatial dependence inherent in the smooth but stochastic spectra.
To account for these modeling challenges, we need a more complex model.

@@

Here we describe our first contribution: the use of a Bayesian hierarchical model 
to estimate the underlying spectra and quantify uncertainty for a particular cosmology,
independent of the others. In the previous seciton, we used lowercase $\bar{y}$ to denote 
the observed weighted average; here, we use $\bar{Y}$ to indicate the corresponding 
random variable within the statistical model. Specifically, our input $X$ is the 
$n$ values of $\log_{10}(k)$ and the output is the weighted average $\bar{Y}$. 
Let $D = \{X, \bar{Y}, \Lambda\}$ represent the aggregated 
data for a particular cosmology.  Let $S$ represent the true (but unknown) underlying
power matter spectra from which $D$ was generated.
\textcolor{red}{(@Annie You suggested removing $\Lambda$ from $D$; is fixed and 
doesn't change here, but it would if $X$ or $Y$ changed. Also, we use $\Lambda$ to pre-scale the data)}

@@

Here, we introduce the Bayesian Gaussian Process (GP) model. We assume 
the matter power spectrum $S$ for a given cosmology follows a GP with mean $\mu$ 
and covariance matrix $\Sigma_S$. From the data, we obtain $\bar Y$, which is a 
noisy version of $S$.

@@

\begin{align}
\bar Y|S &\sim GP\left(S,\Sigma_\varepsilon\right) \\
S &\sim GP\left(\mu, \Sigma_S(x,x')\right)
\end{align}

We transform $X$ values to fall within $[0,1]$ and $\bar Y$ so that the mean and variance 
are 0 and 1, respectively. This allows for a straightforward specification of priors, 
where the hyperparameters for the Mat\'ern covariance $\Sigma_S$ each take a vague prior. 
Specifically:

@@

A key property of GPs is that any finite set of $n$ observations from the GP is a 
multivariate normal distribution. Therefore, upon integrating out $S$, we obtain 
$\bar{Y} \sim \mathcal{N}_n(\mu, \Sigma_S+\Sigma_\varepsilon)$ for the $n$ observed X locations. 
Additionally, this model yields the posterior distribution $S|\bar Y \sim \mathcal{N}_n(m, C)$, 
where $C=\left(\Sigma_S^{-1}+\Sigma_\varepsilon^{-1}\right)^{-1}$ and 
$m=C\left(\Sigma_\varepsilon^{-1}\bar Y+\Sigma_S^{-1}\mu\right)$. 
This distribution provides the posterior mean and corresponding uncertainty for 
$S$ within a given cosmology. 
% With this hierarchical model, we improve the efficiency of estimating $S|\bar{Y}$.
